---
title: "Fantasy Consistency Analysis"
author: "Samuel Ivanecky"
date: "3/10/2022"
output: html_document
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=14, fig.height=8) 

library(tidymodels)
library(httr)
library(dplyr)
library(ggrepel)
library(ggimage)
library(gganimate)
library(plotly)
library(ggthemes)
library(ggpubr)
library(ggimage)
library(reshape2)
library(nflfastR)
library(directlabels)
library(nflreadr)
library(nflplotR)
library(kit)

# Run script for fantasy agg data
source("/Users/z0041ss/Personal Projects/nfl/getNFLData.R")

# Load data
# Load player IDs
ff_ids <- nflreadr::load_ff_playerids() %>%
  select(
    gsis_id, draft_year, draft_round, draft_pick, college
  ) %>%
  filter(!is.na(gsis_id)) %>%
  funique()

# Load fantasy expectations
ff_exp <- nflreadr::load_ff_opportunity(seasons = c(2006:2021))

# Join draft data to fantasy data
ff_exp <- ff_exp %>%
  left_join(ff_ids, by = c("player_id" = "gsis_id")) %>%
  funique()

# Clean up data for fantasy expectations
ff_exp_agg <- ff_exp %>%
  filter(week <= 18) %>%
  group_by(season, player_id, full_name, position) %>%
  summarise(
    n_games = n_distinct(week),
    avg_pts = round(mean(total_fantasy_points, na.rm = T), 3),
    avg_pts_diff = round(mean(total_fantasy_points_diff, na.rm = T), 3),
    avg_pts_exp = round(mean(total_fantasy_points_exp, na.rm = T), 3),
    sd_pts = round(sd(total_fantasy_points, na.rm = T), 3),
    sd_pts_diff = round(sd(total_fantasy_points_diff, na.rm = T), 3),
    sd_pts_exp = round(sd(total_fantasy_points_exp, na.rm = T), 3),
    drft_yr = first(draft_year),
    drft_round = first(draft_round),
    drft_pick = first(draft_pick)
  ) %>%
  ungroup() %>%
  group_by(player_id) %>%
  arrange(season, .by_group = T) %>%
  mutate(
    year_in_league = as.numeric(season) - as.numeric(drft_yr) + 1,
    avg_pts_shift = avg_pts - lag(avg_pts),
    avg_pts_diff_shift = avg_pts_diff - lag(avg_pts_diff),
    avg_pts_exp_shift = avg_pts_exp - lag(avg_pts_exp),
    sd_pts_shift = sd_pts - lag(sd_pts),
    sd_pts_diff_shift = sd_pts_diff - lag(sd_pts_diff),
    sd_pts_exp_shift = sd_pts_exp - lag(sd_pts_exp),
    prev_season_games = lag(n_games)
  )

```

## Get Player Teams
```{r}
plyrTeams <- ff %>%
  group_by(player_id, season) %>%
  filter(week == max(week, na.rm = T)) %>%
  select(
    player_id, recent_team, season
  ) %>%
  funique()
```

## Join Player Teams to Aggregate Data
```{r}
ff_agg <- ff_agg %>%
  left_join(plyrTeams, by = c("player_id", "season")) %>%
  funique()
```

# Analysis
## Year 2 RBs
### Subset Data
```{r}
yr2rb <- ff_exp_agg %>%
  filter(position == "RB" & year_in_league == 2 & prev_season_games > 5 & n_games > 5)
```

### Plot Overall Distribution
```{r}
yr2rb %>%
  ggplot(aes(avg_pts_shift)) +
  geom_histogram(fill = "lightgrey", color = "blue", binwidth = 1) +
  ggplot2::labs(
    x = "Avg. Points Shift from Year 1 to Year 2",
    y = "Count",
    caption = "Data: @nflfastR",
    title = "RB Fantasy Points Average Change from Year 1 to Year 2",
    subtitle = "Players required to have minimum of 5 games in both seasons to qualify."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    # legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```

### Plot Individual Marks
```{r}
yr2rb %>%
  ggplot(aes(avg_pts, avg_pts_shift, color = posteam)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 3, colour = "black", fill = "white", max.overlaps = 5) +
  ggplot2::labs(
    x = "Avg. Fantasy Points in Year 2",
    y = "Avg. Point Shift vs Year 1",
    caption = "Data: @nflfastR",
    title = "RB Fantasy Jumps from Year 1 to Year 2",
    subtitle = "Players from 2006 to 2021. RBs required to have at least 5 games played in both seasons."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```

### Year 1 v Year 2
```{r}
yr2rb %>%
  ggplot(aes((avg_pts - avg_pts_shift), avg_pts, color = posteam, size = 0.5)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 3, colour = "black", fill = "white", max.overlaps = 5) +
  ggplot2::labs(
    x = "Avg. Fantasy Points in Year 1",
    y = "Avg. Fantasy Points in Year 2",
    caption = "Data: @nflfastR",
    title = "RB Fantasy Averages in Year 1 vs Year 2",
    subtitle = "Players from 2006 to 2021. RBs required to have at least 5 games played in both seasons."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```

## Year 2 WRs
### Subset Data
```{r}
yr2wr <- ff_exp_agg %>%
  filter(position == "WR" & year_in_league == 2 & prev_season_games > 5 & n_games > 5)
```

### Year 1 v Year 2
```{r}
yr2wr %>%
  ggplot(aes((avg_pts - avg_pts_shift), avg_pts, color = posteam, size = 0.5)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 3, colour = "black", fill = "white", max.overlaps = 5) +
  ggplot2::labs(
    x = "Avg. Fantasy Points in Year 1",
    y = "Avg. Fantasy Points in Year 2",
    caption = "Data: @nflfastR",
    title = "WR Fantasy Averages in Year 1 vs Year 2",
    subtitle = "Players from 2006 to 2021. WRs required to have at least 5 games played in both seasons."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```

### Yr 1 v Yr 2 Exp
```{r}
yr2wr %>%
  ggplot(aes((avg_pts - avg_pts_shift), avg_pts, color = posteam, size = 0.5)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 3, colour = "black", fill = "white", max.overlaps = 5) +
  ggplot2::labs(
    x = "Avg. Fantasy Points in Year 1",
    y = "Avg. Fantasy Points in Year 2",
    caption = "Data: @nflfastR",
    title = "WR Fantasy Averages in Year 1 vs Year 2",
    subtitle = "Players from 2006 to 2021. WRs required to have at least 5 games played in both seasons."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```

### Classify WRs
```{r}
yr2wr <- yr2wr %>%
  mutate(
    shift_cat = case_when(
      avg_pts_shift < 1.5 & avg_pts_shift > -1.5 ~ 'maintained',
      avg_pts_shift >= 1.5 & avg_pts_shift < 4 ~ 'modest improvement',
      avg_pts_shift >= 4 ~ 'substantial improvement',
      avg_pts_shift <= -1.5 & avg_pts_shift > -4 ~ 'modest regression',
      T ~ 'substantial regression'
    )
  )
```

#### Plot Counts by Category
```{r}
yr2wr %>%
  ggplot(aes(shift_cat, fill = shift_cat)) +
  geom_bar(stat = "count") +
  ggplot2::labs(
    x = "Shift Category",
    y = "Count",
    caption = "Data: @nflfastR",
    title = "WR Performances by Category",
    subtitle = "Players from 2006 to 2021. WRs required to have at least 5 games played in both seasons."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```

#### Scatter w Cats
```{r}
yr2wr %>%
  ggplot(aes((avg_pts - avg_pts_shift), avg_pts, color = shift_cat)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 3, colour = "black", fill = "white", max.overlaps = 5) +
  ggplot2::labs(
    x = "Avg. Fantasy Points in Year 1",
    y = "Avg. Fantasy Points in Year 2",
    caption = "Data: @nflfastR",
    title = "WR Fantasy Averages in Year 1 vs Year 2",
    subtitle = "Players from 2006 to 2021. WRs required to have at least 5 games played in both seasons."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    # legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  )
```
## WR Modeling Attempt
### Get WR Data from Year 1
```{r}
# Subset year 1 data
yr1wr <- ff_exp_agg %>%
  ungroup() %>%
  filter(year_in_league == 1 & position == "WR" & full_name %in% yr2wr$full_name) %>%
  select(-c(year_in_league, avg_pts_shift, avg_pts_diff_shift, avg_pts_exp_shift, sd_pts_diff_shift, sd_pts_exp_shift, sd_pts_diff_shift,
            prev_season_games, sd_pts_shift, drft_yr, position)) 

# Subset cats
wr_cats <- yr2wr %>%
  select(full_name, player_id, shift_cat)

# Join data
yr1wr <- yr1wr %>%
  left_join(wr_cats, by = c("full_name", "player_id"))
```

#### Visualize Counts by Draft Round
```{r}
yr1wr %>%
  filter(!is.na(shift_cat)) %>%
  ggplot(aes(drft_round, fill = shift_cat)) +
  geom_bar(stat = "count", position = "dodge") +
  ggplot2::labs(
    x = "Round Drafted",
    y = "Counts",
    fill = "Category",
    caption = "Data: @nflfastR",
    title = "Fantasy Point Shift for WRs Between Rookie & Sophomore Seasons",
    subtitle = "Shifts based on average points per game in year 1 vs year 2. Minimum 5 games played in each season. Data from 2006-2021."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    # legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold")
  ) +
  scale_fill_brewer(palette="Set1") +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7))
```

### Subset Data & Throw Into Model
```{r}
wr_sub <- yr1wr %>%
  select(-c(season, posteam, player_id))
```

### Model
#### Load Libraries
```{r}
library(rsample)
library(caret)
# library(h2o)
# 
# # h2o set-up 
# h2o.no_progress()  # turn off h2o progress bars
# h2o.init()         # launch h2o
```

#### Split Into Train/Test
```{r}
# Split into train/test data
set.seed(5)  

# Convert vars
wr_sub <- wr_sub %>%
  mutate(
    drft_round = as.factor(drft_round),
    shift_cat = as.factor(shift_cat)
  )

# Create split using stratified sampling for shift_cat
split_pt  <- initial_split(wr_sub, prop = 0.7, strata = "shift_cat")

train  <- training(split_pt)
test   <- testing(split_pt)

# Drop player names
train <- train %>%
  select(-c(full_name))

test <- test %>%
  select(-c(full_name))

# Validate splits are similar between
table(train$shift_cat) %>% prop.table()
table(test$shift_cat) %>% prop.table()
```

#### Load Decision Tree Libraries
```{r}
library(rpart)
library(rpart.plot)
library(vip)
library(pdp)
```

#### Create Model
```{r}
wrTree <- rpart(
  formula = shift_cat ~ .,
  data = train,
  method = "class"
)
```

##### Visualize Tree
```{r}
rpart.plot(wrTree, type = 1) +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 35, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=25, face = "bold"),
  )
```

##### Plot CP
```{r}
plotcp(wrTree) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 35, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=25, face = "bold"),
  )
```

##### Feature Importance
```{r}
vip(wrTree, geom = "point", aesthetics = list(size = 7)) +
  ggplot2::labs(
    x = "Feature",
    y = "Feature Importance",
    caption = "Data: @nflfastR",
    title = "Feature Importance in Base WR Rookie Jump Model",
    subtitle = ""
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 35, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=25, face = "bold"),
  )
```

## WR Analysis of 2021 Shifts
### Aggregate Data
```{r}
wrDf <- ff_exp_agg %>%
  filter(position == "WR") %>%
  group_by(player_id, full_name) %>%
  arrange(season, .by_group = TRUE) %>%
  mutate(
    c_avg_pts = cummean(avg_pts),
    c_avg_pts_diff = cummean(avg_pts_diff),
    c_avg_pts_shift = cummean(avg_pts_shift)
  ) %>%
  filter(!is.na(year_in_league)) %>%
  mutate(
    szn_v_c_avg = avg_pts - lag(c_avg_pts)
  )
```

#### Join in Player Teams
```{r}
wrDf <- wrDf %>%
  left_join(plyrTeams, by = c("player_id", "season"))
```

### Plot
```{r}
wrDf %>%
  filter(season == 2021) %>%
 # filter(abs(szn_v_c_avg) >= 3) %>%
  ggplot(aes(avg_pts, szn_v_c_avg, color = posteam)) +
  geom_point(aes(size = 5)) +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 4, colour = "black", fill = "white", max.overlaps = 7) +
  nflplotR::scale_color_nfl(type = "primary") +
  # nflplotR::scale_fill_nfl(type = "primary")
  ggplot2::labs(
    x = "Avg. PPR Points",
    y = "Season Avg vs Career Avg (PPR Points)",
    caption = "Data: @nflfastR",
    title = "WR 2021 Fantasy Averages vs Career",
    subtitle = "Career averages are up to but not including the 2021 season. Positive values indicate current season was better than career average."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 30, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=15, face = "bold"),
  )
```

## RB Analysis of 2021 Shifts
### Aggregate Data
```{r}
rbDf <- ff_exp_agg %>%
  filter(position == "RB") %>%
  group_by(player_id, full_name) %>%
  arrange(season, .by_group = TRUE) %>%
  mutate(
    c_avg_pts = cummean(avg_pts),
    c_avg_pts_diff = cummean(avg_pts_diff),
    c_avg_pts_shift = cummean(avg_pts_shift)
  ) %>%
  filter(!is.na(year_in_league)) %>%
  mutate(
    szn_v_c_avg = avg_pts - lag(c_avg_pts)
  )
```

#### Join in Player Teams
```{r}
rbDf <- rbDf %>%
  left_join(plyrTeams, by = c("player_id", "season"))
```

### Plot
```{r}
rbDf %>%
  filter(season == 2021 & n_games >= 8) %>%
  # filter(abs(szn_v_c_avg) >= 3) %>%
  ggplot(aes(avg_pts, szn_v_c_avg, color = posteam)) +
  geom_point(aes(size = 5)) +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 4, colour = "black", fill = "white", max.overlaps = 7) +
  nflplotR::scale_color_nfl(type = "primary") +
  # nflplotR::scale_fill_nfl(type = "primary")
  ggplot2::labs(
    x = "Avg. PPR Points",
    y = "Season Avg vs Career Avg (PPR Points)",
    caption = "Data: @nflfastR & ff_opportunity",
    title = "RB 2021 Fantasy Averages vs Career",
    subtitle = "Career averages not including the 2021 season. Positive values indicate current season better than career avg. Min. 8 games."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 30, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=15, face = "bold"),
  )
```

### YOY Trend
```{r}
rb21 <- rbDf %>%
  filter(season == 2021 & n_games >= 8 & szn_v_c_avg <= -3)

rbDf %>%
  filter(player_id %in% rb21$player_id) %>%
  filter(season >= 2017) %>%
  # filter(abs(szn_v_c_avg) >= 3) %>%
  ggplot(aes(season, szn_v_c_avg, color = posteam, group = player_id)) +
  geom_point() +
  geom_line() +
  # geom_hline(yintercept = 0, linetype = "dashed") + 
  # geom_label_repel(aes(label = full_name), nudge_y = -0.1, size = 4, colour = "black", fill = "white", max.overlaps = 7) +
  nflplotR::scale_color_nfl(type = "primary") +
  # nflplotR::scale_fill_nfl(type = "primary")
  ggplot2::labs(
    x = "Avg. PPR Points",
    y = "Season Avg vs Career Avg (PPR Points)",
    caption = "Data: @nflfastR & ff_opportunity",
    title = "RB 2021 Fantasy Averages vs Career",
    subtitle = "Career averages not including the 2021 season. Positive values indicate current season better than career avg. Min. 8 games."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 30, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=15, face = "bold"),
  )
```
