---
title: "Season-to-Date Analysis"
author: "Samuel Ivanecky"
date: "12/17/2021"
output: html_document
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=14, fig.height=8) 

library(tidymodels)
library(httr)
library(dplyr)
library(ggrepel)
library(ggimage)
library(nflfastR)
library(gganimate)
library(plotly)
library(ggthemes)
library(ggpubr)
library(ggimage)
library(reshape2)
library(nflfastR)
library(nflplotR)
library(directlabels)
library(nflreadr)
library(nflplotR)
library(kit)
```

# Run Data Load Scripts
```{r}
# Kick off scripts to load fantasy data
# source("/Users/samivanecky/git/nfl/analysis/fantasy/fpPtsvProjScrapeR.R")
# source("/Users/samivanecky/git/nfl/analysis/fantasy/getNFLData.R")

# Load logos
logos <- teams_colors_logos %>%
  select(team_abbr, team_logo_espn)
```

# Data Manipulation
```{r}
# Add data for rolling three game average and season long average
ffDf <- ff %>%
  group_by(full_name, recent_team) %>%
  arrange(week, .by_group = TRUE) %>%
  mutate( 
    var_as_pct = round(abs((Var / fantasy_points_ppr)) * 100, 2)
  ) %>%
  mutate(
    c_avg_ppr_pts = round(cummean(fantasy_points_ppr), 2),
    c_var_ppr_pts = round(cummean(Var), 2),
    c_var_abs_ppr_pts = round(cummean(abs(Var)), 2),
    c_var_pct = round(cummean(var_as_pct), 2),
    ppr_pts_3wk_avg = round(zoo::rollmean(fantasy_points_ppr, k = 3, fill = NA, align = "right"), 2),
    ppr_pts_3wk_var = round(zoo::rollmean(Var, k = 3, fill = NA, align = "right"), 2),
    abs_var_3wk = round(zoo::rollmean(abs(Var), k = 3, fill = NA, align = "right"), 2),
    var_pct_3wk = round(zoo::rollmean(var_as_pct, k = 3, fill = NA, align = "right"), 2)
  ) %>%
  mutate(
    Var_v_Szn_Avg = Var - lag(c_var_ppr_pts),
    PPR_v_Szn_Avg = fantasy_points_ppr - lag(c_avg_ppr_pts),
    Var_v_3wk_Avg = Var - lag(ppr_pts_3wk_var),
    PPR_v_3wk_Avg = fantasy_points_ppr - lag(ppr_pts_3wk_avg),
    PPR_3wk_v_Szn = ppr_pts_3wk_avg - lag(c_avg_ppr_pts, 3),
    Var_3wk_v_Szn = ppr_pts_3wk_var - lag(c_var_ppr_pts, 3),
    Abs_Var_3wk_v_Szn = abs_var_3wk - lag(c_var_abs_ppr_pts, 3),
    Var_3wk_Pct_v_Szn = var_pct_3wk - lag(c_var_pct, 3)
  ) %>%
  mutate(
    `PPR 3/wk Avg.` = case_when(
      ppr_pts_3wk_avg < 5 ~ "< 5",
      ppr_pts_3wk_avg >= 5 & ppr_pts_3wk_avg <= 10 ~ "5-10",
      ppr_pts_3wk_avg > 10 & ppr_pts_3wk_avg <= 17 ~ "11-17",
      T ~ "18+"
    ),
    `Variance 3/wk Avg. (%)` = case_when(
      var_pct_3wk <= 25 ~ "Stable",
      var_pct_3wk > 25 & var_pct_3wk <= 50 ~ "Moderately Volatile",
      var_pct_3wk > 50 ~ "Very Volatile",
      T ~ "Other"
    ),
    `Player Trend` = case_when(
      abs(lag(PPR_3wk_v_Szn)) <= 3 ~ "Neutral",
      lag(PPR_3wk_v_Szn) < -3 ~ "Negative",
      lag(PPR_3wk_v_Szn) > 3 ~ "Positive",
      T ~ "Not Enough Games"
    )
  ) %>%
  mutate(
    `Player Trend` = as.factor(`Player Trend`)
  )
```

# Visuals
## Points Over Projection Graphs
```{r}
ffDf %>%
  filter(full_name != "AJ Dillon") %>%
  filter(week == 15 & position == "QB" & !is.na(fantasy_points_ppr) & Var > 0 & fantasy_points_ppr >= 18) %>%
  ggplot(aes(reorder(full_name, Var), Var, fill = `Player Trend`)) +
  geom_bar(stat = "identity") +
  geom_label(aes(label = fantasy_points_ppr), nudge_y = -0.1, size = 8, colour = "black", fill = "white") +
  ggplot2::labs(
    x = "Player",
    y = "PPR Performance vs Projection",
    caption = "Data: @nflfastR",
    title = "2021 Fantasy Football QBs in Week 15",
    subtitle = "Data does not include Week 15's Tuesday games. Labels show actual PPR points for the week. Min of 18 points."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    # legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 25, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=12, face = "bold"),
  ) +
  coord_flip() +
  scale_fill_brewer(palette="Set1")
```

## Table Points Over Projection
```{r}
top10 <- ffDf %>%
  ungroup() %>%
  filter(week == 15) %>%
  slice_max(n = 10, order_by = Var) %>%
  select(
    full_name, fantasy_points_ppr, Proj.Pts, Var
  ) %>%
  rename(
    Name = full_name,
    `PPR Points - Scored` = fantasy_points_ppr,
    `PPR Points - Expected` = Proj.Pts,
    `Points Over Projection` = Var
  )

DT::datatable(top10)

bot10 <- ffDf %>%
  ungroup() %>%
  filter(week == 15) %>%
  slice_min(n = 10, order_by = Var) %>%
  select(
    full_name, fantasy_points_ppr, Proj.Pts, Var
  ) %>%
  rename(
    Name = full_name,
    `PPR Points - Scored` = fantasy_points_ppr,
    `PPR Points - Expected` = Proj.Pts,
    `Points Over Projection` = Var
  )

DT::datatable(bot10)

```

# Receiving FP by Throws
## Subset PBP Data
```{r}
pass <- pbp %>%
  filter(!is.na(posteam) & play_type == "pass" & !is.na(receiver_player_id) & complete_pass == 1) %>%
  select(
    game_id, game_date, posteam, week, passer_player_id, passer_player_name, receiver_player_id, receiver_player_name, air_yards, yards_after_catch, yards_gained,
    touchdown, game_seconds_remaining,
  ) %>%
  mutate(
    ppr_pts = 1 + (0.1 * yards_gained) + (6 * touchdown)
  )

# Data validation
passGames <- pass %>%
  group_by(
    receiver_player_id, receiver_player_name, week
  ) %>%
  summarise(
    total_ppr_pts = sum(ppr_pts)
  )

# Join data for validation
tempDf <- ff %>%
  filter(position %in% c("WR")) %>%
  select(
    player_name, player_id, week, fantasy_points_ppr
  )

passGames <- passGames %>%
  inner_join(tempDf, by = c("receiver_player_id" = "player_id", "week")) %>%
  mutate(
    pt_diff = round(fantasy_points_ppr - total_ppr_pts, 2)
  )

# Validate
sum((passGames$pt_diff == 0), na.rm = T) / nrow(passGames)
```

## Break Up Data into Pass Lengths
```{r}
pass <- pass %>%
  mutate(
    pass_length = case_when(
      air_yards <= 5 ~ "short",
      air_yards > 5 & air_yards <= 15 ~ "medium",
      air_yards > 15 ~ "long",
      T ~ "other"
    )
  ) 

# Group data
passGrp <- pass %>%
  group_by(
    receiver_player_id, receiver_player_name, pass_length
  ) %>%
  summarise(
    ppr_pts = sum(ppr_pts, na.rm = T)
  )

# Create season breakout grp
passGrpSznGames <- pass %>%
  group_by(
    receiver_player_id, receiver_player_name, week
  ) %>%
  summarise(
    ppr_pts = sum(ppr_pts, na.rm = T)
  ) %>%
  mutate(
    szn_brk = case_when(
      week <= 8 ~ 'H1',
      T ~ 'H2'
    )
  ) %>%
  ungroup() %>%
  group_by( 
    receiver_player_id, receiver_player_name, szn_brk
  ) %>%
  summarise(
    ppr_pts = sum(ppr_pts, na.rm = T),
    games = n_distinct(week)
  ) %>%
  ungroup() %>%
  mutate(
    pts_per_game = round(ppr_pts / games, 2)
  )


# Subset FF data
plyrData <- ff %>%
  filter(position == "WR") %>%
  select(
    player_id, full_name, recent_team, fantasy_points_ppr
  ) %>%
  funique() %>%
  group_by(
    player_id, full_name, recent_team
  ) %>%
  summarise(
    ppr = sum(fantasy_points_ppr, na.rm = T)
  ) %>%
  ungroup() %>%
  filter(!(full_name %in% c("Ced Wilson", "DJ Moore", "AJ Dillon", "Jamycal Hasty"))) %>%
  group_by(recent_team) %>%
  arrange(desc(ppr), .by_group = T) %>%
  mutate(
    team_rank = 1:n()
  ) %>%
  filter(team_rank <= 4)

# Join
passGrp <- passGrp %>%
  inner_join(plyrData, by = c("receiver_player_id" = "player_id"))

passGrpSznGames <- passGrpSznGames %>%
  inner_join(plyrData, by = c("receiver_player_id" = "player_id"))
```

## Visualize Passing Points
```{r}
passGrp %>%
  filter(pass_length != "other") %>%
  ggplot(aes(reorder(full_name, -team_rank), ppr_pts)) +
  geom_col(aes(color = recent_team, fill = recent_team, alpha = pass_length)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "",
    y = "",
    caption = "Data: @nflfastR",
    title = "2021 Fantasy Football WRs PPR Points by Pass Type",
    subtitle = "Showing top 4 WRs per team. Short: < 5 air yds. Medium: 6-15 air yds. Long: > 15 air yds."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 20, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=10, face = "bold")
  ) +
  guides(alpha=guide_legend(title="Pass Length")) +
  coord_flip() +
  facet_wrap(recent_team ~ ., scales = "free_y")
```

## Visualize Points by Szn Quarter
```{r}
passGrpSznGames %>%
  ggplot(aes(reorder(full_name, -team_rank), ppr_pts)) +
  geom_col(aes(color = recent_team, fill = recent_team, alpha = szn_brk)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "",
    y = "",
    caption = "Data: @nflfastR",
    title = "2021 Fantasy Football WRs Avg. PPR Points per Game by Season Halves",
    subtitle = "Showing top 4 WRs per team. Season segments are broken into four: H1 is Weeks 1-8, H2 is Weeks 9-16. Only showing PPR points for receptions."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 20, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=10, face = "bold")
  ) +
  guides(alpha=guide_legend(title="Season Segment")) +
  coord_flip() +
  facet_wrap(recent_team ~ ., scales = "free_y")
```

# Running Back Points by Run Length
## Aggregate Data
```{r}
run <- pbp %>%
  filter(!is.na(posteam) & play_type == "run" & !is.na(rusher_player_id)) %>%
  select(
    game_id, game_date, posteam, week, rusher_player_id, rusher_player_name, yards_gained, touchdown, game_seconds_remaining
  ) %>%
  mutate(
    ppr_pts = (0.1 * yards_gained) + (6 * touchdown)
  )

# Data validation
runGames <- run %>%
  group_by(
    rusher_player_id, rusher_player_name, week
  ) %>%
  summarise(
    total_ppr_pts = sum(ppr_pts)
  )

# Join data for validation
tempDf <- ff %>%
  filter(position %in% c("RB")) %>%
  select(
    player_name, player_id, week, fantasy_points_ppr
  )

runGames <- runGames %>%
  inner_join(tempDf, by = c("rusher_player_id" = "player_id", "week")) %>%
  mutate(
    pt_diff = round(fantasy_points_ppr - total_ppr_pts, 2)
  )

# Validate
sum((runGames$pt_diff == 0), na.rm = T) / nrow(passGames)
```

## Break Up Data into Rush Lengths
```{r}
run <- run %>%
  mutate(
    rush_length = case_when(
      yards_gained <= 5 ~ "short",
      yards_gained > 5 & yards_gained <= 10 ~ "medium",
      yards_gained > 10 ~ "long",
      T ~ "other"
    )
  ) 

# Group data
runGrp <- run %>%
  group_by(
    rusher_player_id, rusher_player_name, rush_length
  ) %>%
  summarise(
    ppr_pts = sum(ppr_pts, na.rm = T)
  )

# Subset FF data
plyrData <- ff %>%
  filter(position == "RB") %>%
  select(
    player_id, full_name, recent_team, fantasy_points_ppr
  ) %>%
  funique() %>%
  group_by(
    player_id, full_name, recent_team
  ) %>%
  summarise(
    ppr = sum(fantasy_points_ppr, na.rm = T)
  ) %>%
  ungroup() %>%
  filter(!(full_name %in% c("Ced Wilson", "DJ Moore", "AJ Dillon"))) %>%
  group_by(recent_team) %>%
  arrange(desc(ppr), .by_group = T) %>%
  mutate(
    team_rank = 1:n()
  ) %>%
  filter(team_rank <= 3)

# Join
runGrp <- runGrp %>%
  inner_join(plyrData, by = c("rusher_player_id" = "player_id"))
```

## Visualize Rushing Points
```{r}
runGrp %>%
  filter(rush_length != "other") %>%
  ggplot(aes(reorder(full_name, ppr_pts), ppr_pts)) +
  geom_col(aes(color = recent_team, fill = recent_team, alpha = rush_length)) +
  # geom_label(aes(label = full_name), nudge_y = -0.1, size = 8, colour = "black", fill = "white") +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "",
    y = "",
    caption = "Data: @nflfastR",
    title = "2021 Fantasy Football RBs Points by Rush Length",
    subtitle = "Showing top 3 RBs per team. Short: < 5 yds. Medium: 6-10 yds. Long: > 10 yds. Data through Week 15."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    # legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 45, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=20, face = "bold")
  ) +
  guides(alpha=guide_legend(title="Rush Length")) +
  coord_flip() +
  facet_wrap(recent_team ~ ., scales = "free_y")
```

# Skill Player Points by Play Length
## Aggregate Data
```{r}
# Get player data to join for position
plyrInfo <- ff %>%
  filter(position %in% c("WR", "RB", "TE")) %>%
  group_by(player_id) %>%
  summarise(
    position = max(position, na.rm =T)
  ) %>%
  funique()
  
  
plays <- pbp %>%
  filter(!is.na(posteam) & ((play_type == "run" & !is.na(rusher_player_id)) | (play_type == "pass" & !is.na(receiver_player_id)))) %>%
  mutate(
    player_id = coalesce(receiver_player_id, rusher_player_id),
    player_name = coalesce(receiver_player_name, rusher_player_name)
  ) %>%
  select(
    game_id, game_date, posteam, week, player_id, player_name, yards_gained, air_yards, touchdown, game_seconds_remaining
  ) %>%
  mutate(
    ppr_pts = (0.1 * yards_gained) + (6 * touchdown)
  ) %>%
  left_join(
    plyrInfo, by = c("player_id")
  )
```

## Break Up Data into Play Lengths
```{r}
plays <- plays %>%
  mutate(
    play_length = case_when(
      yards_gained <= 5 & position == "RB" ~ "short",
      yards_gained > 5 & yards_gained <= 10 & position == "RB" ~ "medium",
      yards_gained > 10 & position == "RB" ~ "long",
      yards_gained <= 5 & position %in% c("WR", "TE") ~ "short",
      yards_gained > 5 & air_yards <= 15 & position %in% c("WR", "TE") ~ "medium",
      yards_gained > 15 & position %in% c("WR", "TE") ~ "long",
      T ~ "other"
    )
  ) 

# Group data
playsGrp <- plays %>%
  group_by(
    player_id, player_name, play_length
  ) %>%
  summarise(
    ppr_pts = sum(ppr_pts, na.rm = T)
  )

# Subset FF data
plyrData <- ff %>%
  filter(position %in% c("RB", "TE", "WR")) %>%
  select(
    player_id, full_name, recent_team, fantasy_points_ppr
  ) %>%
  funique() %>%
  group_by(
    player_id, full_name, recent_team
  ) %>%
  summarise(
    ppr = sum(fantasy_points_ppr, na.rm = T)
  ) %>%
  ungroup() %>%
  filter(!(full_name %in% c("Ced Wilson", "DJ Moore", "AJ Dillon"))) %>%
  group_by(recent_team) %>%
  arrange(desc(ppr), .by_group = T) %>%
  mutate(
    team_rank = 1:n()
  ) %>%
  filter(team_rank <= 4)

# Join
playsGrp <- playsGrp %>%
  inner_join(plyrData, by = c("player_id"))
```

## Visualize Play Points
```{r}
playsGrp %>%
  filter(play_length != "other") %>%
  ggplot(aes(reorder(full_name, ppr_pts), ppr_pts)) +
  geom_col(aes(color = recent_team, fill = recent_team, alpha = play_length)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "",
    y = "",
    caption = "Data: @nflfastR",
    title = "2021 Fantasy Football RBs Points by Rush Length",
    subtitle = "Showing top 3 RBs per team. Short: < 5 yds. Medium: 6-10 yds. Long: > 10 yds. Data through Week 15."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    # legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 20, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=10, face = "bold")
  ) +
  guides(alpha=guide_legend(title="Play Length")) +
  coord_flip() +
  facet_wrap(recent_team ~ ., scales = "free_y")
```

# QB EPA by Pass Length
## Get Data
```{r}
qbEpa <- pbp %>%
  filter()
```



