---
title: "WR Breakout Seasons"
author: "Samuel Ivanecky"
date: "4/8/2022"
output: html_document
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=14, fig.height=8) 

library(tidymodels)
library(httr)
library(dplyr)
library(ggrepel)
library(gganimate)
library(plotly)
library(ggthemes)
library(ggpubr)
library(ggimage)
library(reshape2)
library(nflfastR)
library(directlabels)
library(nflreadr)
library(nflplotR)
library(kit)
library(cluster)  
library(factoextra)
library(gt)
library(reactable)
library(htmltools)

# Run script for fantasy agg data
source("/Users/z0041ss/Personal Projects/nfl/getNFLData.R")

# Load data
# Load player IDs
ff_ids <- nflreadr::load_ff_playerids() %>%
  select(
    gsis_id, draft_year, draft_round, draft_pick, college
  ) %>%
  filter(!is.na(gsis_id)) %>%
  funique()

# Load fantasy expectations
ff_exp <- nflreadr::load_ff_opportunity(seasons = c(2006:2021))

# Join draft data to fantasy data
ff_exp <- ff_exp %>%
  left_join(ff_ids, by = c("player_id" = "gsis_id")) %>%
  funique()

# Clean up data for fantasy expectations
ff_exp_agg <- ff_exp %>%
  filter(week <= 18) %>%
  group_by(season, player_id, full_name, position) %>%
  summarise(
    n_games = n_distinct(week),
    avg_pts = round(mean(total_fantasy_points, na.rm = T), 3),
    avg_pts_diff = round(mean(total_fantasy_points_diff, na.rm = T), 3),
    avg_pts_exp = round(mean(total_fantasy_points_exp, na.rm = T), 3),
    sd_pts = round(sd(total_fantasy_points, na.rm = T), 3),
    sd_pts_diff = round(sd(total_fantasy_points_diff, na.rm = T), 3),
    sd_pts_exp = round(sd(total_fantasy_points_exp, na.rm = T), 3),
    drft_yr = first(draft_year),
    drft_round = first(draft_round),
    drft_pick = first(draft_pick)
  ) %>%
  ungroup() %>%
  group_by(player_id) %>%
  arrange(season, .by_group = T) %>%
  mutate(
    year_in_league = as.numeric(season) - as.numeric(drft_yr) + 1,
    avg_pts_shift = avg_pts - lag(avg_pts),
    avg_pts_diff_shift = avg_pts_diff - lag(avg_pts_diff),
    avg_pts_exp_shift = avg_pts_exp - lag(avg_pts_exp),
    sd_pts_shift = sd_pts - lag(sd_pts),
    sd_pts_diff_shift = sd_pts_diff - lag(sd_pts_diff),
    sd_pts_exp_shift = sd_pts_exp - lag(sd_pts_exp),
    prev_season_games = lag(n_games)
  )

```

## Get Player Teams
```{r}
plyrTeams <- ff %>%
  group_by(player_id, season) %>%
  filter(week == max(week, na.rm = T)) %>%
  select(
    player_id, recent_team, season
  ) %>%
  funique()
```

## Subset Player Names
```{r}
plyrNames <- ff %>%
  select(
    player_id, season, full_name
  ) %>%
  group_by(
    player_id, season
  ) %>%
  summarise(
    full_name = first(full_name)
  ) %>%
  ungroup() %>%
  funique()
```

## Join Player Teams to Aggregate Data
```{r}
ff_agg <- ff_agg %>%
  left_join(plyrTeams, by = c("player_id", "season")) %>%
  funique()
```

# 2021 Wide Receivers
## Subset Data
```{r}
# Get players from 2021 minimum 3 games
ff21_plyrs <- ff_agg %>%
  filter(season == 2021 & n_games >= 3)

# Subset seasons for players from 2021 season w at least 5 games
ff21 <- ff_agg %>%
  filter(player_id %in% ff21_plyrs$player_id) %>%
  filter(n_games >= 5)

# Subset WRs
wr21 <- ff21 %>%
  filter(position == "WR")

# Subset RBs
rb21 <- ff21 %>%
  filter(position == "RB")
```

## New Metrics
### Subset 2021 WRs
```{r}
# Create a season change metric for average points
wr21 <- wr21 %>%
  group_by(player_id) %>%
  arrange(season, .by_group = TRUE) %>%
  mutate(
    szn_avg_change = avg_pts_ppr - lag(avg_pts_ppr)
  ) %>%
  mutate(
    szn_avg_change_pct = round((szn_avg_change / avg_pts_ppr) * 100, 3)
  ) %>%
  ungroup()
```

## WR Breakout Subset Visuals
```{r}
# Looking at WR who increased production by minimum 30%
brkoutWr <- wr21 %>%
  filter(szn_avg_change_pct >= 25 & avg_pts_ppr >= 9 & n_games >= 10)

# Subset data to these receivers
brkOutDf <- wr21 %>%
  filter(player_id %in% brkoutWr$player_id) %>%
  mutate(
    is_brkout_szn = case_when(
      szn_avg_change_pct >= 25 & avg_pts_ppr >= 10 & n_games >= 10 ~ 'Y',
      T ~ 'N'
    )
  ) %>%
  group_by(player_id) %>%
  arrange(season, .by_group = TRUE) %>%
  mutate(
    is_post_brkout = case_when(
      lag(is_brkout_szn) == 'Y' ~ 'Y',
      T ~ 'N'
    )
  )

# Look at average change 
brkOutDf %>%
  filter(is_post_brkout == "Y") %>%
  mutate(
    plyr_name_grp = paste0(player_name, " - ", season)
  ) %>%
  ggplot(aes(reorder(plyr_name_grp, szn_avg_change), szn_avg_change)) +
  geom_col(aes(color = recent_team, fill = recent_team)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "Shift in Avg. PPR Points vs Breakout Season",
    y = "",
    caption = "Data: @nflfastR",
    title = "Fantasy Breakout Season Continued Success - Wide Receivers",
    subtitle = "Showing shift in season average PPR points for the season following a breakout year. Breakout year identified by >= 25% improvement from previous season."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 45, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=20, face = "bold")
  ) +
  coord_flip()

# Look at average change 
brkOutDf %>%
  filter(is_post_brkout == "Y") %>%
  mutate(
    plyr_name_grp = paste0(player_name, " - ", season)
  ) %>%
  ggplot(aes(reorder(plyr_name_grp, szn_avg_change_pct), szn_avg_change_pct)) +
  geom_col(aes(color = recent_team, fill = recent_team)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "Shift in Avg. PPR Points vs Breakout Season",
    y = "",
    caption = "Data: @nflfastR",
    title = "Fantasy Breakout Season Continued Success - Wide Receivers",
    subtitle = "Showing shift in season average PPR points for the season following a breakout year. Breakout year identified by >= 25% improvement from previous season."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 45, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=20, face = "bold")
  ) +
  coord_flip()

# Look at average change 
brkOutDf %>%
  filter(is_post_brkout == "Y" & season == 2021) %>%
  mutate(
    plyr_name_grp = paste0(player_name, " - ", season)
  ) %>%
  ggplot(aes(reorder(plyr_name_grp, szn_avg_change), szn_avg_change)) +
  geom_col(aes(color = recent_team, fill = recent_team)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "Shift in Avg. PPR Points vs Breakout Season",
    y = "",
    caption = "Data: @nflfastR",
    title = "2020 Breakout WRs in 2021",
    subtitle = "Showing shift in season average PPR points for the season following a breakout year. Breakout year identified by >= 25% improvement from previous season."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 45, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=20, face = "bold")
  ) +
  coord_flip()

brkOutDf %>%
  filter(is_post_brkout == "Y" & season == 2021) %>%
  mutate(
    plyr_name_grp = paste0(player_name, " - ", season)
  ) %>%
  ggplot(aes(reorder(plyr_name_grp, szn_avg_change_pct), szn_avg_change_pct)) +
  geom_col(aes(color = recent_team, fill = recent_team)) +
  nflplotR::scale_color_nfl(type = "secondary") +
  nflplotR::scale_fill_nfl(type = "primary") +
  ggplot2::labs(
    x = "Shift in Avg. PPR Points vs Breakout Season",
    y = "",
    caption = "Data: @nflfastR",
    title = "2020 Breakout WRs in 2021",
    subtitle = "Showing shift in season average PPR points for the season following a breakout year. Breakout year identified by >= 25% improvement from previous season."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 45, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=20, face = "bold")
  ) +
  coord_flip()
```

### Table Values
#### Create Metric
```{r}
postBrkOutSzns <- brkOutDf %>%
  filter(is_post_brkout == "Y") %>%
  mutate(
    pct_change_cat = case_when(
      szn_avg_change_pct >= 10 ~ "Improved",
      szn_avg_change_pct < 10 & szn_avg_change_pct >= -5 ~ "Maintained",
      szn_avg_change_pct < -5 & szn_avg_change_pct >= -15 ~ "Moderate Regression",
      T ~ "Major Regression"
    )
  )
```

#### Plot Data
```{r}
postBrkOutSzns %>%
  ggplot(aes(pct_change_cat, fill = pct_change_cat)) +
  geom_bar(stat = "count") +
  ggplot2::labs(
    x = "Performance Category Following Breakout Season",
    y = "Number of Receivers",
    caption = "Data: @nflfastR. Data for WRs active in 2021.",
    title = "Fantasy Breakout Season Continued Success - Wide Receivers",
    subtitle = "Showing player breakout by performance in PPR season average following breakout season."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    legend.position = "none",
    axis.text.y = element_text(face = "italic"),
    plot.title = ggplot2::element_text(size = 45, hjust = 0.5, vjust = 0.5, face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    text = element_text(size=20, face = "bold")
  ) 
```

#### Players Who Maintained/Improved
```{r}
# Subset players
mainImp <- postBrkOutSzns %>%
  filter(
    pct_change_cat %in% c("Maintained", "Improved")
  )
```

### Who Broke Out in 2021?
#### Data
```{r}
# View table
brkOutDf %>%
  filter(season == 2021 & is_brkout_szn == "Y") %>%
  select(
    full_name, n_games, avg_pts_ppr, szn_avg_change
  ) %>%
  rename(
    Name = full_name,
    `Games Played` = n_games,
    `Avg. Points PPR` = avg_pts_ppr,
    `Season Avg. Improvement vs Last Year` = szn_avg_change
  ) %>%
  ungroup() %>%
  select(-c(player_id)) %>%
  reactable()
```

### Examine Previous Breakout WRs
#### Define Pre-Breakout Szn
```{r}
brkOutDf <- brkOutDf %>%
  group_by(player_id) %>%
  arrange(season, .by_group = TRUE) %>%
  mutate(
    is_pre_brkout_szn = case_when(
      lead(is_brkout_szn) == "Y" ~ "Y",
      T ~ "N"
    )
  ) %>%
  ungroup()
```

#### Subset Pre-Breakout Szns
```{r}
# Subset season stats for WRs
wr_stats <- season_stats %>%
  filter(season_type == "REG") %>%
  select(
    player_id, season, carries, rushing_yards, rushing_tds, rushing_epa, receptions, targets, receiving_yards, receiving_tds,
    receiving_air_yards, receiving_yards_after_catch, receiving_epa, target_share, air_yards_share
  )

# Subset and join stats from that season
preBrk <- brkOutDf %>%
  filter(is_pre_brkout_szn == "Y") %>%
  left_join(wr_stats, by = c("player_id", "season")) %>%
  mutate(
    target_share = round((target_share / n_games) * 100, 4),
    air_yards_share = round((air_yards_share / n_games) * 100, 4)
  )
```

#### Clustering Analysis
##### Check for NA Data
```{r}
# Check numeric columns for NA values
preBrk %>%
  summarise(
    across(where(is.numeric), ~ sum(is.na(.x)))
  ) 
```

##### Subset Data for Clustering
```{r}
preBrk <- preBrk %>%
  mutate(
    lbl = paste0(full_name, "-", season)
  ) %>%
  select(-c(player_name, position, season, szn_avg_change, szn_avg_change_pct, is_brkout_szn, is_post_brkout, 
            is_pre_brkout_szn, avg_pt_pct, avg_pt_pct_ppr, player_id, full_name))

# Set row names
rownames(preBrk) <- preBrk$lbl
```

##### Normalize Data
```{r}
# Scale data
preBrk <- preBrk %>%
  mutate_if(is.numeric, scale)

# Reformat row names
preBrk <- preBrk %>%
  remove_rownames() %>%
  column_to_rownames(var = "lbl")
```

##### Check Distance
```{r}
distance <- get_dist(preBrk)
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))
```

##### K Means
```{r}
temp <- preBrk %>%
  select(-c(recent_team))

k3 <- kmeans(temp, centers = 5, nstart = 25)
str(k3)

fviz_cluster(k3, data = temp)
```

### Predict Breakout Szn
#### Combine Data
```{r}
# Subset and join stats from that season
wr <- brkOutDf %>%
  left_join(wr_stats, by = c("player_id", "season")) %>%
  filter(season > 2008) %>%
  mutate(
    target_share = round((target_share / n_games) * 100, 4),
    air_yards_share = round((air_yards_share / n_games) * 100, 4)
  ) %>%
  na.omit() %>%
  ungroup() %>%
  mutate(
    lbl = paste0(full_name, "-", season, "-", recent_team)
  ) %>%
  funique() %>%
  remove_rownames() %>%
  column_to_rownames(var = "lbl")
```

## WR Clustering
### Subset Data
```{r}
# Subset and join stats from that season
wr <- wr %>%
  left_join(wr_stats, by = c("player_id", "season", "recent_team")) %>%
  filter(season > 2005) %>%
  mutate(
    target_share = round((target_share / n_games) * 100, 4),
    air_yards_share = round((air_yards_share / n_games) * 100, 4)
  ) %>%
  na.omit() %>%
  ungroup() %>%
  mutate(
    lbl = paste0(full_name, "-", season, "-", recent_team)
  ) %>%
  filter(full_name != "Cordarrelle Patterson") %>%
  select(-c(player_name, position, season, avg_pt_pct, avg_pt_pct_ppr, player_id, full_name)) %>%
  distinct(lbl, .keep_all = TRUE) 
 
# Scale data
wr <- wr %>%
  mutate_if(is.numeric, scale)

# Reformat row names
wr <- wr %>%
  remove_rownames() %>%
  column_to_rownames(var = "lbl")
```

#### Cluster
```{r}
temp <- wr %>%
  select(-c(recent_team, abs_med_pt_pct))

k3 <- kmeans(temp, centers = 10, nstart = 25)
str(k3)

fviz_cluster(k3, data = temp)
```

